---
title: "Fatal Force Data Visualization"
output: html_document
---
Objective: Analyze the Washington Post's fatal force data set, with a focus on data visualization. 

(Link to data set: https://github.com/washingtonpost/data-police-shootings)

Load packages and download data set.
```{r}

library(tidyverse)
library(readr)
library(lubridate)
library(knitr)
library(kableExtra)
library(tinytex)
setwd("~/Desktop/RStudio Directory")
police_data <- read_csv('fatal-police-shootings-data (2).csv')

```


Review data set. 
```{r}

view(police_data)

```


Determine the first and last dates on which shootings occurred and the total number of fatal shootings that occurred over this time frame.
```{r}

head(police_data)

tail(police_data)

nrow(police_data)

```


- First recorded shooting: January 2, 2015
- Last recorded shooting: December 10, 2021
- Total fatal shootings: 6,800


Determine how many fatal shootings occurred per year from 2015 - 2021.
```{r}

police_data_tally <- police_data %>% mutate(assign_1 = 1)

police_data_year_col <- police_data_tally %>% 
                        mutate(year = year(police_data_tally$date))

table_value_1 <- police_data_year_col %>% group_by(year) %>% 
                                          summarize(yearly_sums = sum(assign_1))

table_value_1 %>% 
  kbl(col.names = c('Year', 'Total Fatal Shootings Per Year'), align = c('c', 'c')) %>%
  kable_classic_2(full_width = F)

```


Determine the average number of fatal shootings per year from 2015 - 2021. (Note: the data set is missing shooting data for the last couple weeks of December 2021, so the average listed below is likely less than the true average.)
```{r}

yearly_sums_data <- police_data_year_col %>% 
                    group_by(year) %>%
                    summarize(yearly_sums = sum(assign_1)) %>% 
                    mutate(yearly_sums = yearly_sums)

yearly_sums_data %>% summarize(mean(yearly_sums)) %>% 
                     mutate_if(is.numeric, round, digits = 2)

```


From the code above we see that approximately 971 fatal shootings occurred per year for the years on record. 

Create a bar graph that shows the total number of fatal shootings per year from 2015 - 2021. 
```{r}

ggplot() + 
geom_col (yearly_sums_data, mapping = aes(x = year, y = yearly_sums, fill = as.factor(year))) + 
geom_text(yearly_sums_data, mapping = aes(x = year, y = yearly_sums, label = yearly_sums, vjust = -.2)) +
scale_x_discrete(limits = c(2015, 2016, 2017, 2018, 2019, 2020, 2021),
                 labels = c("2015", "2016", 
                            "2017", "2018", 
                            "2019", "2020", "2021")) +
                 labs(title = "Total Fatal Shootings Per Year", 
                      x = "Year", 
                      y = "Fatal Shootings per Year", 
                      fill = 'Year') +
                 theme(plot.title = element_text(hjust = 0.5))

```


Create line graphs that show fatal shooting rates on a month-by-month basis for each year.

(Note: there is a steep drop off in the 2021 graph because of the missing values in December referenced above.)
```{r}

monthly_freq <- police_data %>% mutate(year = year(date), month = month(date)) %>%
                                group_by(year, month) %>%
                                summarise(n = n())

monthly_freq %>% ggplot(aes(x = month, y = n, color = year)) +
                 geom_point(monthly_freq, mapping = aes(color = as.factor(year))) +
                 geom_line(monthly_freq, mapping = aes(color = as.factor(year))) +
                 facet_wrap(vars(year)) +
                 scale_x_discrete(limits = c(2, 4, 6, 8, 10, 12),
                 labels = c("2", "4", "6", "8", "10", "12")) +
                 scale_y_discrete(limits = c(20, 40, 60, 80, 100, 120),
                 labels = c("20", "40", "60", "80", "100", "120")) +
                 labs(title = "Total Fatal Shootings per Month", 
                      x = "Months", 
                      y = "Fatal Shootings per Month",
                      fill = 'Year') +
                 theme(plot.title = element_text(hjust = 0.5))


```


Determine the mean, median, and range with respect to age for individuals who were fatally shot. 
```{r}

police_data %>% summarize(mean(age, na.rm = T),
                          median(age, na.rm = T),
                          range(age, na.rm = T))

left <- c(37.13, ' ')
middle <- c(35, ' ')
right <- c(6, 92)

table_value_2 <- data.frame (left, middle, right)


table_value_2 %>% 
kbl(col.names = c('Mean', 'Median', 'Range'), align = c('c', 'c', 'c')) %>%
kable_classic_2(full_width = F)

```


Create a histogram that shows the age distribution of individuals who were fatally shot.
```{r}

ggplot(police_data, aes(x = age)) + 
geom_histogram (aes(y =..density..), 
                breaks = seq(0, 100, by = 1), 
                colour = "black", 
                fill = "green") +
stat_function(fun = dnorm, 
              args = list(mean = mean(police_data$age), 
              sd = sd(police_data$age))) +
geom_density(bins = 100) +
labs(title = "Age Distribution of Individuals Who Were Fatally Shot", 
     x = "Age", 
     y = "Density") +
theme(plot.title = element_text(hjust = 0.5))

```


Compare the number of males who were fatally shot to the number of females who were fatally shot. 
```{r}

gender_breakdown <- police_data %>% 
                    group_by(gender) %>%
                    drop_na() %>% 
                    count() %>% 
                    ungroup() %>% 
                    mutate(perc = `n` / sum(`n`)) %>% 
                    arrange(perc) %>%
                    mutate(labels = scales::percent(perc)) %>% 
                    mutate_if(is.numeric, round, digits = 4)

gender_breakdown %>% kbl(col.names = c('Gender', 
                                       'Total Fatal Shootings', 
                                       'Decimal', 
                                       'Rounded Percentage'), 
                         align = c('c', 'c', 'c', 'c')) %>%
                         kable_classic_2(full_width = F)

```


Create a pie chart that compares the number of males who were fatally shot to the number of females who were fatally shot. 
```{r}

gender_breakdown %>% ggplot(aes(x= "", y = perc, fill = gender)) +
                     geom_bar(stat="identity", width=1) +
                     geom_text(aes(label = labels),
                                   position = position_stack(vjust = 0.52), 
                                   cex = 2.9) +
                     coord_polar("y", start=0) +
                     labs(title = "Fatalities Broken Down by Gender",
                          x = ' ' ,
                          y = "Percentage",
                          fill = 'Gender') +
                     theme(plot.title = element_text(hjust = 0.5))

```


Determine the average age of men and women who were fatally shot.
```{r}

table_value_4 <- police_data %>% 
                 drop_na() %>% 
                 group_by(gender) %>% 
                 summarize(mean(age)) %>% 
                 mutate_if(is.numeric, round, digits = 3)

table_value_4 %>% 
  kbl(col.names = c('Gender', 'Average Age When Fatally Shot'), 
      align = c('c', 'c')) %>%
  kable_classic_2(full_width = F)

```


Create a box and whisker plot that incorporates data on age and gender. 
```{r}

police_data %>% drop_na() %>% 
                ggplot(mapping = aes(x = gender, y = age, fill = gender)) + 
                geom_boxplot() + 
                labs(title = "Age Distribution Between Sexes",
                          x = 'Gender of Individual Shot' ,
                          y = 'Age', 
                          fill = 'Gender') +
                theme(plot.title = element_text(hjust = 0.5))

```

Determine the breakdown of race among individuals who were fatally shot. 
```{r}

race_breakdown <- police_data %>% 
                  mutate(race_string = 
                         case_when(race == 'W' ~ 'White, non-Hispanic',
                                   race == 'B' ~ 'Black, non-Hispanic',
                                   race == 'A' ~ 'Asian',
                                   race == 'N' ~ 'Native American',
                                   race == 'H' ~ 'Hispanic',
                                   race == 'O' ~ 'Other'))


table_value_5 <- race_breakdown %>% group_by(race_string) %>%
                 drop_na() %>% 
                 count() %>% 
                 ungroup() %>% 
                 mutate(perc = `n` / sum(`n`)) %>% 
                 arrange(perc) %>%
                 mutate(rounded = scales::percent(perc)) %>% 
                 mutate_if(is.numeric, round, digits = 4) %>% 
                 arrange(desc(n))


table_value_5 %>% kbl(col.names = c('Race', 
                                    'Total Fatal Shootings', 
                                    'Decimal', 
                                    'Unrounded Percentage'), 
                      align = c('c', 'c')) %>%
                      kable_classic_2(full_width = F)

```


Create a bar graph that shows the shooting breakdown by race. 
```{r}

ggplot() + 
geom_col (table_value_5, mapping = aes(x = race_string, 
                                       y = perc, fill = as.factor(race_string))) + 
geom_text(table_value_5, mapping = aes(x = race_string, 
                                       y = perc, label = perc, 
                                       vjust = -.2)) +
scale_x_discrete(limits = c("Asian", "Black, non-Hispanic", 
                            "Hispanic", "Native American", 
                            "Other", "White, non-Hispanic"),
                 labels = c("A", "B", 
                            "H", "N", 
                            "O", "W")) +
labs(title = "Breakdown of Fatal Shootings by Race", 
     x = "Race", 
     y = "Proportions", 
     fill = 'Race') +
theme(plot.title = element_text(hjust = 0.5))

```


Determine the average age among individuals who were shot, according to race.
```{r}

table_value_6 <- race_breakdown %>% 
                 drop_na() %>% 
                 group_by(race_string) %>% 
                 summarize(mean_age = mean(age)) %>% 
                 mutate_if(is.numeric, round, digits = 3) %>% 
                 arrange(desc(mean_age))

table_value_6 %>% kbl(col.names = c('Race', 'Average Age When Fatally Shot'), 
                      align = c('l', 'c')) %>% kable_classic_2(full_width = F)

```


Create a box and whisker plot that incorporates data on age and race.
```{r}

race_breakdown %>% 
drop_na() %>% 
ggplot(mapping = aes(x = race_string, y = age, fill = race_string)) + 
geom_boxplot() + 
scale_x_discrete(limits = c("Asian","Black, non-Hispanic", "Hispanic", 
                              "Native American", "Other", "White, non-Hispanic"),
                   labels = c("A", "B","H", "N", "O", "W")) +
                   labs(title = "Age Distribution Among Races",
                        x = 'Race of Individual Shot' ,
                        y = 'Age', 
                        fill = 'Race') +
theme(plot.title = element_text(hjust = 0.5))   

```


A few more interesting tables.

Five states with the most shootings from 2015 - 2021.
```{r}

top_5 <- police_data %>% 
                  drop_na() %>% 
                  group_by(state) %>% 
                  summarize(totals = n()) %>% 
                  mutate(frequency = totals / sum(totals)) %>% 
                  mutate_at(vars(starts_with("frequency")), funs(round(., 4))) %>% 
                  arrange(desc(totals)) %>% 
                  head(5)

top_5 %>% 
kbl(col.names = c('Top States', 'Total Fatal Shootings', 'Proportions'), 
    align = c('l', 'c')) %>% 
kable_classic_2(full_width = F)

```


Five states with the least shootings from 2015 - 2021.
```{r}

bottom_5 <- police_data %>% 
                  drop_na() %>% 
                  group_by(state) %>% 
                  summarize(totals = n()) %>% 
                  mutate(frequency = totals / sum(totals)) %>% 
                  mutate_at(vars(starts_with("frequency")), funs(round(., 4))) %>% 
                  arrange(totals) %>% 
                  head(5)

bottom_5 %>% 
kbl(col.names = c('Bottom States', 'Total Fatal Shootings', 'Proportions'), 
    align = c('l', 'c')) %>% 
kable_classic_2(full_width = F)

```


Breakdown of shootings that occurred when individuals who were shot were armed vs. unarmed. 
```{r}

armed_vs_unarmed <- police_data %>% drop_na(armed) %>% 
                                    mutate(armed_vs_unarmed = ifelse (
                                    armed == 'unarmed', 'Unarmed', 'Armed')) %>% 
                                    group_by(armed_vs_unarmed) %>%
                                    summarize (n = n()) %>% 
                                    mutate(freq = n / sum(n)) %>% 
                                    mutate_at(vars(starts_with("freq")), funs(round(., 4)))

     
armed_vs_unarmed %>% 
kbl(col.names = c('Armed vs. Unarmed', 'Total Fatal Shootings', 'Proportions'), 
    align = c('l', 'c')) %>% 
kable_classic_2(full_width = F)

```


Breakdown of shootings that occurred with and without a police body camera.  
```{r}

body_camera <- police_data %>% 
               group_by(body_camera) %>%
               summarize (n = n()) %>% 
               mutate(freq = n / sum(n)) %>% 
               mutate_at(vars(starts_with("freq")), funs(round(., 4)))

     
body_camera %>% 
kbl(col.names = c('Wearing Body Camera', 'Total Fatal Shootings', 'Proportions'), 
    align = c('l', 'c')) %>% 
kable_classic_2(full_width = F)

```

